---
title: "Untitled"
author: "Viderp"
date: "2021/9/4"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Machine Learning assignment Weida Pan(369868) & Zhe Zhao(433707)
R language logistic regression (Logistic regression) model classification prediction patient coronary heart disease risk data analysis report

```{r}
library(tidyverse)
library(broom)
library(pROC)
library(plotROC) 
library(knitr)

```
## Load data 
```{r}
setwd("C:/Users/ICBC/Desktop/exam requirment/machine")
heart_data <- read_csv("framingham.csv") %>%
  drop_na() %>% 
  mutate(education = case_when(
    education == 1 ~ "Some HS", 
    education == 2 ~ "HS or GED", 
    education == 3 ~ "Some College", 
    education == 4 ~ "College"
  ),
  TenYearCHD = as.factor(TenYearCHD), 
  currentSmoker = as.factor(currentSmoker),
  ageCent = age - mean(age), 
  totCholCent = totChol - mean(totChol), 
  BPMeds = as.factor(BPMeds), 
  prevalentStroke = as.factor(prevalentStroke), 
  prevalentHyp = as.factor(prevalentHyp), 
  male = as.factor(male), 
  diabetes = as.factor(diabetes)
  )

```


##Fit logistic regression model
```{r car}
risk_m <- glm(TenYearCHD ~ ageCent + currentSmoker + totCholCent, 
              data = heart_data, family = binomial)
tidy(risk_m, conf.int = TRUE, exponentiate = FALSE) %>% 
  kable(format = "markdown", digits = 3)
```
##prediction for new patients
```{r}
x0 <- data_frame(ageCent = (60 - 49.552), 
                 totCholCent = (263 - 236.848), 
                 currentSmoker = as.factor(0))

x0
```

#Predicted probability

```{r}
predict(risk_m, x0, type = "response")

risk_m_aug <- augment(risk_m, type.predict = "response" )
risk_m_aug
```

##Confusion matrix
```{r}
threshold <- 0.1
risk_m_aug %>%
  mutate(risk_predict = if_else(.fitted > threshold, "1: Yes", "0: No")) %>%
  group_by(TenYearCHD, risk_predict) %>%
  summarise(n = n()) %>%
  kable(format="markdown")


risk_m_aug %>%
  mutate(risk_predict = if_else(.fitted > threshold, "1: Yes", "0: No")) %>%
  select(TenYearCHD, risk_predict)
```

##Roc Curve
```{r}
(roc_curve <- ggplot(risk_m_aug, 
                     aes(d = as.numeric(TenYearCHD) - 1, 
                         m = .fitted)) +
    geom_roc(n.cuts = 10, labelround = 3) + 
    geom_abline(intercept = 0) + 
    labs(x = "False positive rate (1-specificity)", 
         y = "True positive rate (sensitivity)") )

calc_auc(roc_curve)$AUC
```

#Plot the original residuals
```{r}
ggplot(data = risk_m_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predictive value", y = "Original residual")
```

#classify residual plot
```{r}
arm::binnedplot(x = risk_m_aug$.fitted, y = risk_m_aug$.resid,
                xlab = "Predicted probability", 
                main = "Comparison of residual value and predicted 
                value after classification", 
                col.int = FALSE)
```

#The relationship between graded residuals and age
```{r}
arm::binnedplot(x = risk_m_aug$ageCent, 
                y = risk_m_aug$.resid, 
                col.int = FALSE,
                xlab = "Age (centered on the average)", 
                main = "The relationship between graded 
residuals and age")
```

#Classified Residuals and Total Cholesterol
```{r}
arm::binnedplot(x = risk_m_aug$totCholCent, 
                y = risk_m_aug$.resid, 
                col.int = FALSE,
                xlab = "Total cholesterol (centered on average)", 
     
                           main = "Classified Residuals and Total Cholesterol")
```

# Use stepwise regression to select the model
```{r}
full_model <- glm(TenYearCHD ~ ., data = heart_data, family = "binomial")
select_model <- step(full_model, direction = "backward")


tidy(select_model, conf.int = TRUE, exponentiate = FALSE) %>%
kable(format = "markdown", digits = 3)
```

